---
title: "Course Project 1"
author: "Matthew Lau"
date: "December 2, 2017"
output: html_document
---

##Load required packges


```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
```

## 1.1) Download data (Optional if using dataset on GitHUb)


```{r cars}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
filename <- "activity.zip"
download.file(url, destfile=filename)
unzip(zipfile=filename, overwrite = TRUE)
```

## 1.2) Read Data and format date

Load the data, set NA's
```{r}
dataset <- read.table("activity.csv", header=TRUE, na.strings="NA", sep=",")
dataset$date <- ymd(dataset$date)
str(dataset)
```

## 2) Histogram of the total number of steps taken each day

Plot histogram
```{r}
dataset_dailysteps <- dataset %>% 
        group_by(date) %>% 
        summarise(daily_steps = sum(steps), na.rm=TRUE)

ggplot(dataset_dailysteps, aes(x = daily_steps)) +
  geom_histogram(fill = "steelblue") +
  labs(title = "Frequency of Daily Steps", x = "Total Daily Steps per Day", y = "Occurance")
```

## 3) Mean and median number of steps taken each day

```{r}
dataset_summary <- dataset %>% 
        group_by(date) %>% 
        summarize(daily_steps = sum(steps, na.rm=TRUE)) %>% #calculate daily steps
        summarize(mean = mean(daily_steps, na.rm=TRUE), median = median(daily_steps, na.rm=TRUE)) #find average and median
dataset_summary
```


## 4) Time series plot of the average number of steps taken

```{r}
dataset_timeseries <- dataset %>% 
        group_by(date) %>% 
        summarize(daily_steps = sum(steps, na.rm=TRUE))

ggplot(dataset_timeseries, aes(x = date, y = daily_steps)) +
        geom_point(colour = "steelblue", na.rm=TRUE) + #scatterplot
        scale_x_date(breaks=date_breaks("1 week"), labels=date_format("%b %d"))+ #set date scale
        labs(title = "Time Series Plot of Daily Steps from October 2012 to November 2012", x = "Date", y = "Daily Steps")+ #add in labels
        theme(plot.title = element_text(hjust = 0.5)) # center title

```

## 5) The 5-minute interval that, on average, contains the maximum number of steps

### 1) Find the average steps of intervals
```{r}
dataset_interval <- dataset %>% 
        group_by(interval) %>% 
        summarize(interval_average_steps = mean(steps, na.rm=TRUE))
head(dataset_interval, 10)

ggplot(dataset_interval, aes(x = interval, y = interval_average_steps)) +
        geom_line(colour = "steelblue", na.rm=TRUE) +
        labs(title = "Plot of Average Steps by Interval", x = "Interval", y = "Daily Steps")+
        theme(plot.title = element_text(hjust = 0.5))

```

### 2) Find the interval with highest steps

```{r}
dataset_highest_interval <- dataset_interval %>% 
        arrange(-interval_average_steps) %>% 
        head(1)
dataset_highest_interval
```


## 6) Code to describe and show a strategy for imputing missing data

### 1) Calculate # of missing values
```{r}
dataset_totalrows <- nrow(dataset) 
dataset_completedcases <- sum(complete.cases(dataset))
dataset_NACases <- dataset_totalrows - dataset_completedcases
dataset_NACases

```

### 2) Strategy for imputting missing data
Use the mean of the interval to calculate missing data

```{r}
dataset_impute <- dataset %>% 
        left_join(dataset_interval, by=c("interval"="interval")) %>%
        mutate(steps = case_when(
                is.na(steps) == TRUE ~ as.integer(round(interval_average_steps, 0)),
                TRUE ~ as.integer(steps))) %>%
        select (-interval_average_steps)
head(dataset_impute, 10)

NACases <- sum(is.na(dataset_impute))
NACases
```

### 3) Mean and Media of new imputed dataset

Yes, the imputed values differ from original dataset. It has increased the mean and the median.

```{r}
dataset_impute_summary <- dataset_impute %>% 
        group_by(date) %>% 
        summarize(daily_steps = sum(steps, na.rm=TRUE)) %>% #calculate daily steps
        summarize(mean = mean(daily_steps, na.rm=TRUE), median = median(daily_steps, na.rm=TRUE)) #find average and median
dataset_impute_summary
```

## 7) Histogram of the total number of steps taken each day after missing values are imputed

```{r}
dataset_dailysteps_impute <- dataset_impute %>% 
        group_by(date) %>% 
        summarise(daily_steps = sum(steps, na.rm=TRUE))

ggplot(dataset_dailysteps_impute, aes(x = daily_steps)) +
        geom_histogram(fill = "steelblue") +
        labs(title = "Frequency of Daily Steps", x = "Total Daily Steps per Day", y = "Occurance")+
        theme(plot.title = element_text(hjust = 0.5))

```


## 8) Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
dataset_dayofweek_impute <- dataset_impute %>% 
        mutate(dayofweek = wday(date, label=TRUE),
               weekend = ifelse(dayofweek %in% c("Sun", "Sat"), "Weekend", "Weekday ")) %>%
        group_by(weekend, interval) %>% 
        summarise(daily_steps = sum(steps, na.rm=TRUE))

ggplot(dataset_dayofweek_impute, aes(x = interval, y = daily_steps)) +
        geom_line(colour = "steelblue") +
        facet_grid(weekend ~ .) +
        labs(title = "Time Series Plot of Daily Steps on Weekdays vs Weekend", x = "Interval", y = "Total Steps")+
        theme(plot.title = element_text(hjust = 0.5))

```
